Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2022-03-15 doc/peakcharacterization_manuscript.Rmd
general improvements to the manuscript

---
title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
author: "Robert M Flight, Joshua M Mitchell, Hunter NB Moseley"
date: "`r Sys.time()`"
output: 
  word_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

@@

To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [citation]. 
The sliding and tiled windows are also converted to IRanges objects the same way.

@@

  1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
  2. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
  3. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
  4. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
  5. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
  6. Noise removal, two pass normalization, frequency SD filtering (filtersd)

@@

Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly.
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan number; 2 - Some peak differences between scans are correlated with intensity.
The solution to **1** is to do a two pass normalization.
In the first pass, the peaks who's intensity is correlated with intensity are detected.
In the second pass, the correlated peaks are removed, and normalization is carried out again.
Figure XA contains an example peak whose intensity across scans is correlated with scan number.
The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in Figure XB.
The highlighted peaks in Figure XB are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in Figure XA and XB.

```{r increment_intensity_scan}
figure_count$increment("intensity_scan")
```

@@

```{r normalization_factors, eval = FALSE}
tar_load(normalization_plot_normalization_values_data)
normalization_plot_normalization_values_data
```

*caption*: **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity >= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm). 
**B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.

@@

```{r rsd_table, results = 'asis'}
tar_load(rsd_values)
knitr::kable(rsd_values, digits = 3)
```

@@

>>>>

---
title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
author:
  - Robert M Flight:
      institute: [markey, biochem, rcsirm]
  - Joshua M Mitchell:
      institute: lab
  - Hunter NB Moseley:
      email: hunter.moseley@uky.edu
      correspondence: true
      institute: [markey, biochem, rcsirm, ibi, tox]
institute:
  - markey: Markey Cancer Center, University of Kentucky, Lexington, KY 40536, USA
  - biochem: Department of Molecular & Cellular Biochemistry, University of Kentucky, Lexington, KY 40536, USA
  - rcsirm: Resource Center for Stable Isotope Resolved Metabolomics, University of Kentucky, Lexington, KY 40536, USA
  - ibi: Institute for Biomedical Informatics, University of Kentucky, Lexington, KY 40536, USA
  - tox: Department of Toxicology and Cancer Biology, University of Kentucky, Lexington, KY 40536, USA
  - lab: Somewhere over the rainbow
date: "`r Sys.time()`"
output: 
  word_document:
    keep_md: true
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: '`r here::here("doc/peakcharacterization.json")`'
csl: plos-computational-biology.csl
editor_options: 
  chunk_output_type: console
---

@@

To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a]. 
The sliding and tiled windows are also converted to IRanges objects the same way.

@@

  1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
  1. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
  1. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
  1. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
  1. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
  1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
  1. Scan level peak centroids generated by MSnbase
  1. Scans merged and then centroids generated by MSnbase

For the two sets of MSnbase peaks, we also matched scan level peaks to merged peaks via two methods:

  1. All scan level peaks within a 2ppm window are associated with the *merged* peak
  1. Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop M/Z of the *filtersd* peaks

@@

Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly.
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan number; 2 - Some peak differences between scans are correlated with intensity.
The solution to **1** is to do a two pass normalization.
In the first pass, the peaks who's intensity is correlated with scan are detected.
In the second pass, the correlated peaks are removed, and normalization is carried out again.
Figure XA contains an example peak whose intensity across scans is correlated with scan number.
The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in Figure XB.
The highlighted peaks in Figure XB are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in Figure XA and XB.

```{r increment_intensity_scan}
figure_count$increment("intensity_scan")
figure_count$increment("normalization_factors")
```

@@

```{r normalization_factors, dn_id = figure_count, fig.height = 8, fig.width = 8}
tar_load(compare_normalization)
compare_normalization[[1]]
```

`r figure_count$label_text("normalization_factors")`. 
**A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity >= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm). 
**B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.

@@

```{r rsd_table, results = 'asis'}
tar_load(rsd_values)
rsd_ft = flextable(rsd_values) %>%
  colformat_double(digits = 2) %>%
  autofit()
rsd_ft
```

@@

## References
